### 2020-05-30 writeup/intro.tex
intro

We combine all the labeled data using a hard parameter sharing architecture that contains a shared body $B\in\real^{p\times r}$ for all tasks and a separate prediction head $\set{W_i \in \real^{r}}_{i=1}^t$ for each task \cite{R17,MTDNN19,WZR20}.
This corresponds to minimizing the following objective.
\begin{align}
	\label{eq_mtl}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2.
\end{align}
For a target task $t$,
let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal multi-task estimator by minimizing equation \eqref{eq_mtl}.
Let $\hat{\beta}_t^{\STL}$ denote the standard linear regression estimator using task $t$ alone.
We say there is negative transfer if the test error of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.

@@

To quantify the tradeoff more precisely, we develop a new technical result that provides a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, for the setting of two tasks with general covariance matrices.
Our result extends a well-known result on the trace of $(X_1^{\top}X_1)^{-1}$ for a single random matrix \cite{S07}.
Furthermore, we provide a nearly optimal error bound, which may be of independent interest.

Using the result, we provide a sharp analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for the setting of two tasks.
For more than two tasks, we show a similar result if all tasks have the same features.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
Finally, we apply our technical tools to the related setting of transfer learning.
We study the transfer function used by Taskonomy \cite{ZSSGM18}, which pools learnt representations from source tasks into a shared body similar to $B$ in equation \eqref{eq_mtl}.
We find that the model shift bias can be captured by the orthogonal projection of $\beta_t$ to the subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
These results are presented more precisely in Section \ref{sec_main}.

The tradeoff that we develop in Section \ref{sec_main} depends on the properties of each task such as its data size and covariance matrix.
In Section \ref{sec_insight}, we use the technical results to provide guidance on when positive transfer is likely to occur and identify causes of negative transfer.
\squishlist
		\item First, we provide a more precise interpretation to a folklore understanding in multi-task learning, which posits that negative transfer is caused by having dissimilar tasks.
		For an example of two tasks, we show a sharp transition from positive to negative transfer determined by the ratio of $\norm{\beta_1 - \beta_2}^2$ and a certain function of data sizes (cf. Proposition \ref{prop_dist_transition}).
		We further show that positive transfer is more likely to occur when transferring from a less noisy source task to a more noisy target task.
		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
		\item Second, we find that depending on how large $\norm{\beta_1 - \beta_2}^2$ is, adding more labeled data from the source task does not always reduce the test error of $\hat{\beta}_t^{\MTL}$ (cf. Proposition \ref{prop_data_size}).
	We connect the observation to explain a key finding of Taskonomy \cite{ZSSGM18}.
	We define the \textit{data efficiency ratio} as the smallest $\alpha\in(0, 1)$ such that if we only use an $\alpha$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.
	In Section \ref{sec_validate}, we validate that performing multi-task learning can reduce the need for labeled data on 6 sentiment analysis tasks. %, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
		\item Finally, we find that covariate shift, i.e. having different covariance matrices, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
		We show that as $n_1 / n_2$ becomes large, the best performing source task has have the same covariance matrix as the target task (cf. Proposition \ref{prop_covariate}).
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.
\squishend

%	A crucial technical tool that we develop for showing Theorem \ref{thm_main_informal} is the asymptotic limit of the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$ (for the setting of two tasks), which may be of independent interest.

We show practical implications of our theory on text and image classification tasks.
	First, we provide a metric to determine positive versus negative transfer by comparing the test accuracies of single-task models.
	Second, \todo{}
	Finally, we show that as the data size between the source and target task becomes more imbalanced, aligning the covariances of the tasks becomes more beneficial.
% as the data size becomes more imbalanced between the source and target task, having mis-aligned task covariance matrices can result in suboptimal performance.
	%validate on text and image classification tasks that comparing single-task accuracies can help determine whether multi-task learning performs better than single-task learning.

%	On the other hand, if the number of source task datapoints is comparable to the target task, aligning task covariances may hurt performance.

>>>>

We combine all the labeled data using a hard parameter sharing architecture that contains a shared body $B\in\real^{p\times r}$ for all tasks and a separate prediction head $\set{W_i \in \real^{r}}_{i=1}^t$ for each task \cite{R17,MTDNN19,WZR20}.
This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
For a target task $t$,
let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal multi-task estimator by minimizing equation \eqref{eq_mtl}.
Let $\hat{\beta}_t^{\STL}$ denote the standard linear regression estimator using task $t$ alone.
We say there is negative transfer if the test error of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.

@@

To quantify the tradeoff more precisely, we develop a new technical tool that provides a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, for the setting of two tasks with general covariance matrices.
Our result extends a well-known result on the trace of $(X_1^{\top}X_1)^{-1}$ for a single random matrix \cite{S07}.
Furthermore, we provide a nearly optimal error bound, which may be of independent interest.
Using the result, we provide a sharp analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two settings:
(i) two tasks with general covariance matrice; (ii) any number of tasks that have the same features (e.g. images).
Finally, we apply our technical tool to the related setting of transfer learning.
We pool learnt representations from source tasks into a shared body similar to $B$ in equation \eqref{eq_mtl}, which has also been used in Taskonomy \cite{ZSSGM18}.
For this setting, we show that the model shift bias can be captured by the orthogonal projection of $\beta_t$ to the subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
These results are presented more precisely in Section \ref{sec_main}.

The tradeoff that we develop depends on the properties of each task such as its data size and covariance matrix.
In Section \ref{sec_insight}, we use our technical tool to rigorously explain many phenomena in multi-task learning.
We consider three factors including task dissimilarity, data size and covariate shift.
\squishlist
		\item First, we define task dissimilarity as the distance between $\beta_1$ and $\beta_2$ using a simplified isotropic setting of two tasks.
		Using our technical tool, we show a sharp transition from positive to negative transfer as task dissimilarity increases.
		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
		\item Second, we consider the effect of uneven data sizes.
	We begin by observing that adding more labeled data from the source task does not always improve performance.
	Then, we theoretically analyze the benefit of multit-task learning for reducing the amount of labeled data needed to achieve comparable performance to single-task learning, which is a key empirical finding of Taskonomy \cite{ZSSGM18}.
%	We define the \textit{data efficiency ratio} as the smallest $x$ such that if we only use an $$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	%In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.
	%In Section \ref{sec_validate}, we validate that performing multi-task learning can reduce the need for labeled data on 6 sentiment analysis tasks. %, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
		\item Finally, we show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
		We show that as $n_1 / n_2$ becomes large, the source task should have the same covariance matrix as the target task to ensure the best performance.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.
\squishend
We validate the above phenomena on text and image classification tasks.
Based on the theory, we provide algorithmic implications for detecting and mitigating negative transfer.
First, we provide a metric to determine positive versus negative transfer by comparing the test accuracies of single-task models.
Second, \todo{}
Finally, we provide a new insight on the covariance alignment algorithm, proposed in \cite{WZR20}.
We show that as the size of the source task becomes larger compared to the target task, the covariance alignment algorithm produces more significant results.

### 2020-05-30 writeup/setup.tex
intro

To illustrate our intuition, we begin by considering the setting of two tasks with general covariance matrices.
Recall that $\hat{\beta}_t^{\MTL}$ is defined as $BW_t$ after solving equation \eqref{eq_mtl}.
We decompose the test error of $\hat{\beta}_{t}^{\MTL}$ on the target task into two parts (to be derived in Appendix \ref{app_proof_sec3}) as follows
\begin{align}
	\te(\hat{\beta}_t^{\MTL}) =& ~ \hat{v}^2 \bignorm{\Sigma_2^{1/2} (\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}X_1^{\top}X_1 (\beta_1 - \hat{v}\beta_2)}^2 \label{eq_te_model_shift} \\
	&+ \sigma^2\cdot \bigtr{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2}, \label{eq_te_var}
\end{align}
where $\hat{v} = W_2 / W_1$ denotes the ratio of the output layer weights.
The scaling term $\hat{v}$ corresponds to the intuition that the shared module $B$ learns a subspace for both tasks.
The output layer of each task scales the direction of $B$ suitably to fit the task.

>>>>

To illustrate our intuition, we begin by considering the setting of two tasks with general covariance matrices.
Recall that $\hat{\beta}_t^{\MTL}$ is defined as $BW_t$ after solving equation \eqref{eq_mtl}.
\begin{align}
	\label{eq_mtl}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2.
\end{align}
We decompose the test error of $\hat{\beta}_{t}^{\MTL}$ on the target task into two parts (to be derived in Appendix \ref{app_proof_sec3}) as follows
\begin{align}
	\te(\hat{\beta}_t^{\MTL}) =& ~ \hat{v}^2 \bignorm{\Sigma_2^{1/2} (\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}X_1^{\top}X_1 (\beta_1 - \hat{v}\beta_2)}^2 \label{eq_te_model_shift} \\
	&+ \sigma^2\cdot \bigtr{(\hat{v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2}, \label{eq_te_var}
\end{align}
where $\hat{v} = W_2 / W_1$ denotes the ratio of the output layer weights.
The scaling term $\hat{v}$ corresponds to the intuition that the shared module $B$ learns a subspace for both tasks.
The output layer of each task scales the direction of $B$ suitably to fit the task.
